---
title: Live Recording of a Midi Controller via Mido Inport
author: Urs Wilke
date: '2020-10-24'
slug: []
categories: []
tags:
  - midi
  - reticulate
subtitle: ''
summary: ''
authors: []
lastmod: '2020-10-24T23:10:55+02:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: true
projects: []
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

## motivation

I was made aware of the idea to import live data of a midi controller through this blog post:

```{r echo=FALSE, eval=TRUE}
blogdown::shortcode('tweet', '1306695952175169539')
```

## mido controller input via mido


To run this on your machine, you need to install the R package [reticulate](https://rstudio.github.io/reticulate/#installation) and the python package [mido](https://mido.readthedocs.io/en/latest/installing.html).


### import mido

```{r load_mido}
reticulate::use_python("/home/chief/anaconda3/bin/python", required = T)

mido <- reticulate::import("mido")
```

### connect to inport

You need to connect your midi controller to the computer. In case you don't have one and want to try this out you can also run a virtual midi controller, for example [VMPK](https://sourceforge.net/projects/vmpk/). In this example I connected a Nektar Impact LX 61.

```{r connect}
# In order to obtain all the inports available, just type
# mido$get_input_names()
# in your console
inport  <- mido$open_input("Impact LX61+:Impact LX61+ MIDI 1 20:0")
```

### live record midi controller

Now we are all set up to record the midi messages coming in via the input port.

We just create empty vectors and append the midi messages and the times when they arrive in an endless loop until we send a keyboard interrupt:

```{r read_data}
msgs <- c()
times <- c()
while (TRUE) {
  t <- Sys.time()
  msg <- inport$poll()
  # only record the midi message when it's not empty:
  if (!is.null(msg)) {
    msgs <<- append(msgs, as.character(msg))
    times <<- append(times, t)
  }
}
```

*Instead of the* `poll()` *method mido also [provides](https://mido.readthedocs.io/en/latest/ports.html#input-ports) 
a* `receive()` *method which should be more appropriate for this action. However, I wasn't able to get this working without making my PC stuck and having to restart R.*

### print out the recorded data

After interrupting the loop above the data is in an R vector msgs. Here I show a small example of the resulting data when a C note is played twice and the pitch wheel is turned a bit:

```{r print_data}
# Set time precision value for printing to milliseconds:
op <- options(digits.secs = 3)
# Print out the recorded data:
data.frame(times, msgs)
#>                     times                                          msgs
#> 1 2020-10-24 14:04:31.611 note_on channel=15 note=60 velocity=81 time=0
#> 2 2020-10-24 14:04:31.742 note_off channel=15 note=60 velocity=0 time=0
#> 3 2020-10-24 14:04:31.981 note_on channel=15 note=60 velocity=78 time=0
#> 4 2020-10-24 14:04:32.117 note_off channel=15 note=60 velocity=0 time=0
#> 5 2020-10-24 14:04:33.927        pitchwheel channel=15 pitch=128 time=0
#> 6 2020-10-24 14:04:33.938        pitchwheel channel=15 pitch=384 time=0
#> 7 2020-10-24 14:04:33.960        pitchwheel channel=15 pitch=512 time=0
#> 8 2020-10-24 14:04:33.992        pitchwheel channel=15 pitch=128 time=0
#> 9 2020-10-24 14:04:34.003          pitchwheel channel=15 pitch=0 time=0

```

